---
title: "title"
author: "Luca Renz & Kasper Aleksander Kuznik"
date: "December 2024"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
    theme: simplex
---

```{r libraries, cache=TRUE}
# Load necessary libraries
#Install and import libraries.
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("lubridate")) install.packages("lubridate")
if (!require("vars")) install.packages("vars")
if (!require("forecast")) install.packages("forecast")
if (!require("rugarch")) install.packages("rugarch")
if (!require("dplyr")) install.packages("dplyr")
if (!require("purrr")) install.packages("purrr")
if (!require("readxl")) install.packages("readxl")
if (!require("janitor")) install.packages("janitor")

library(tidyverse)
library(dplyr)
library(lubridate)
library(vars)
library(forecast)
library(rugarch)
library(purrr)
library(readxl)
library(janitor)  # For cleaning column names
```

## Title
```{r setup, include=FALSE}
# Global options
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```


### Loading data
```{r data-prep, cache=TRUE}
coal_futures <- read.csv("data/NewCastle_Coal_Futures_historical_prices.csv")
elec_generation <- readxl::read_excel("data/US_Elec_Generation_Monthly.xlsx")
all_sheets <- excel_sheets("data/US_Elec_Generation_Monthly.xlsx")
oil_futures <- read.csv("data/Crude_Oil_WTI_Futures_historical_prices.csv")
gas_futures <- read.csv("data/Natural_Gas_Futures_historical_prices.csv")


# Coal Futures
coal_futures <- coal_futures %>%
  mutate(YearMonth = floor_date(as.Date(Exp.Date, format = "%Y-%m-%d"), "month")) %>%
  dplyr::select(YearMonth, Last) %>%
  rename(Coal_Price = Last)

# Oil Futures
oil_futures <- oil_futures %>%
    mutate(YearMonth = floor_date(as.Date(Exp.Date, format = "%Y-%m-%d"), "month")) %>%
  dplyr::select(YearMonth, Last) %>%
  rename(Oil_Price = Last)

# Natural Gas Futures
gas_futures <- gas_futures %>%
    mutate(YearMonth = floor_date(as.Date(Exp.Date, format = "%d/%m/%Y"), "month")) %>%
  dplyr::select(YearMonth, Last) %>%
  rename(Gas_Price = Last)




# Function to remove all single quotes from character columns
remove_quotes <- function(df) {
  df %>%
    mutate(across(where(is.character), ~ gsub("'", "", .))) %>%
     mutate(across(where(is.character), ~ gsub("\\n|\\r", " ", .)))  # Remove line breaks
}

valid_sources <- c("Coal", "Gas", "Oil")
# not as wide table
elec_generation_processed <- all_sheets %>%
  lapply(function(sheet) {
    read_excel("data/US_Elec_Generation_Monthly.xlsx", sheet = sheet) %>%
      remove_quotes() %>%
      clean_names() %>%
      mutate(
  energy_source = case_when(
    energy_source == "Natural gas" ~ "Gas",
    energy_source == "Other Gases" ~ "Gas",
    energy_source == "Petroleum" ~ "Oil",
    TRUE ~ energy_source
  )
) %>%
     filter(state == "US-Total") %>% 
      filter(startsWith(type_of_producer, "Total")) %>% 
     filter(energy_source %in% valid_sources) %>%
      mutate(
        year_month = make_date(year, month)  # Create a YearMonth column
      ) %>%
      dplyr::select(year_month, energy_source, mwh) %>%
      rename(YearMonth = year_month)
  }) %>%
  bind_rows()  # Combine all processed data from different sheets

sum(is.na(coal_futures$YearMonth))
sum(is.na(gas_futures$YearMonth))
sum(is.na(oil_futures$YearMonth))


wide_table <- elec_generation_processed %>%
  pivot_wider(
    names_from = energy_source,  # Use 'energy_source' for column names
    values_from = mwh,          # Use 'mwh' for the values in those columns          # Fill missing values with 0
  )


# Merge all datasets
merged_data <- purrr::reduce(
  list(coal_futures, oil_futures, gas_futures, wide_table),
  dplyr::left_join,
  by = "YearMonth"
)


write.csv(merged_data,'data/merged_data.csv', row.names=FALSE)
```

# Data Analysis
## Exploratory Data Analysis
```{r analysis, cache=TRUE}
# Data filtering from 2012-01-01 to 2024-08-01
merged_data_reduced <- merged_data[38:189,]
# Futures prices
Nat_Gas_Futures_price <- ts(merged_data_reduced$Gas_Price, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
Oil_Futures_price <- ts(merged_data_reduced$Oil_Price, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
Coal_Futures_price <- ts(merged_data_reduced$Coal_Price, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
# US Electricity
Nat_Gas_US_Elec <- ts(merged_data_reduced$Gas, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
Oil_US_Elec <- ts(merged_data_reduced$Oil, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
Coal_US_Elec <- ts(merged_data_reduced$Coal, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)

# Plotting the data
# Futures data
plot(Nat_Gas_Futures_price, xlab = "Time", ylab = "Natural Gas Futures Price [tsd USD]", main = "Natural Gas Future Price January 2012 - August 2024")
plot(Oil_Futures_price, xlab = "Time", ylab = "Oil Futures Price [USD]", main = "Oil Future Price January 2012 - August 2024")
plot(Coal_Futures_price, xlab = "Time", ylab = "Coal Futures Price [USD]", main = "Coal Future Price January 2012 - August 2024")
# US Electricity data
plot(Nat_Gas_US_Elec, xlab = "Time", ylab = "US Electricity Generation - Natural Gas", main = "US Electricity Generation - Natural Gas - January 2012 - August 2024")
plot(Oil_US_Elec, xlab = "Time", ylab = "US Electricity Generation - Oil", main = "US Electricity Generation - Oil - January 2012 - August 2024")
plot(Coal_US_Elec, xlab = "Time", ylab = "US Electricity Generation - Coal", main = "US Electricity Generation - Oil - January 2012 - August 2024")
```

## Testing data for stationarity
```{r }
# Testing for stationarity for both time series and creating stationary time series of same lengths
# Futures data
adf.test(Nat_Gas_Futures_price) # Test for stationarity of the Natural Gas Future price series
adf.test(Oil_Futures_price) # Test for stationarity of the Oil Futures price series
adf.test(Coal_Futures_price) # Test for stationarity of the Coal Futures price series

# US Electricity data
adf.test(Nat_Gas_US_Elec) # Test for stationarity of Nat Gas US Elec Gen Monthly series
adf.test(Oil_US_Elec) # Test for stationarity of Oil US Elec Gen Monthly series
adf.test(Coal_US_Elec) # Test for stationarity of Coal US Elec Monthly generation series

# Apply diff-log-transformation if necessary
# Futures Data
Nat_Gas_Futures_returns <- diff(log(Nat_Gas_Futures_price))
Oil_Futures_returns <- diff(log(Oil_Futures_price))
Coal_Futures_returns <- diff(log(Coal_Futures_price))

# US Electricity Data
Nat_Gas_US_Elec_diff <- diff(log(Nat_Gas_US_Elec))

# Plotting the differences
plot(Nat_Gas_Futures_returns)
plot(Oil_Futures_returns)
plot(Coal_Futures_returns)
plot(Nat_Gas_US_Elec_diff)

# Check for stationarity again
adf.test(Nat_Gas_Futures_returns) # Test for stationarity of Diff of Nat Gas US Elec Gen Monthly series
adf.test(Oil_Futures_returns) # Test for stationarity of Diff of Oil US Elec Gen Monthly series
adf.test(Coal_Futures_returns) # Test for stationarity of Diff of Coal US Elec Monthly generation series
adf.test(Nat_Gas_US_Elec_diff) # Test for stationarity of Diff of Coal US Elec Monthly generation series
```

# 1a. Natural Gas Futures Returns vs Natural Gas US Electricity Generation Difference
```{r}
# Estimating vector autoregession and testing for causality
VAR_est <- VAR(cbind(Nat_Gas_Futures_returns, Nat_Gas_US_Elec_diff), ic="AIC", lag.max = 24)
coeftest(VAR_est)
causality(VAR_est, cause="Nat_Gas_Futures_returns")["Granger"] # Testing for Granger-causal effects of Natural Gas US Electricity Generation on Natural Gas Futures Returns
causality(VAR_est, cause="Nat_Gas_US_Elec_diff")["Granger"] # Testing for Granger-causal effects of Natural Gas Futures Returns on Natural Gas US Electricity Generation

# Interpretation
# TBD

# Degression: Plotting impulse response functions
plot(irf(VAR_est, impulse="Nat_Gas_US_Elec_diff", response="Nat_Gas_Futures_returns")) # We shock Nat_Gas_Futures_price_diff at time t
plot(irf(VAR_est, impulse="Nat_Gas_Futures_returns", response="Nat_Gas_US_Elec_diff")) # We shock Nat_Gas_US_Elec_Diff at time t
# Interpretation:
# TBD

```

Natural Gas Data is not stationary but give p-value 0.06 checking the results out of curiosity
# 1b. Natural Gas Futures Returns vs Natural Gas US Electricity Generation
```{r}
# Estimating vector autoregession and testing for causality
Nat_Gas_US_Elec <- Nat_Gas_US_Elec[-1]
VAR_est <- VAR(cbind(Nat_Gas_Futures_returns, Nat_Gas_US_Elec), ic="AIC", lag.max = 24)
coeftest(VAR_est)
causality(VAR_est, cause="Nat_Gas_Futures_returns")["Granger"] # Testing for Granger-causal effects of Natural Gas US Electricity Generation on Natural Gas Futures Returns
causality(VAR_est, cause="Nat_Gas_US_Elec")["Granger"] # Testing for Granger-causal effects of Natural Gas Futures Returns on Natural Gas US Electricity Generation

# Interpretation
# TBD

# Degression: Plotting impulse response functions
plot(irf(VAR_est, impulse="Nat_Gas_US_Elec", response="Nat_Gas_Futures_returns")) # We shock Nat_Gas_Futures_price_diff at time t
plot(irf(VAR_est, impulse="Nat_Gas_Futures_returns", response="Nat_Gas_US_Elec")) # We shock Nat_Gas_US_Elec at time t
# Interpretation:
# TBD

```

# 2. Oil Futures Returns vs Oil US Electricity Generation
```{r}
Oil_US_Elec <- Oil_US_Elec[-1]
VAR_est <- VAR(cbind(Oil_Futures_price_diff, Oil_US_Elec), ic="AIC", lag.max = 24)
coeftest(VAR_est)
causality(VAR_est, cause="Oil_Futures_returns")["Granger"] # Testing for Granger-causal effects of the US Oil Electricity Generation on oil futures returns
causality(VAR_est, cause="Oil_US_Elec")["Granger"] # Testing for Granger-causal effects of oil futures returns on US Oil Electricity Generation

# Interpretation
# TBD

# Degression: Plotting impulse response functions
plot(irf(VAR_est, impulse="Oil_US_Elec", response="Oil_Futures_returns")) # We shock Oil futures returns at time t
plot(irf(VAR_est, impulse="Oil_Futures_returns", response="Oil_US_Elec")) # We shock US Oil Electricity Generation at time t
# Interpretation:
# TBD

```

# 3. Coal Futures Returns vs Coal US Electricity Generation
```{r}
Coal_US_Elec <- Coal_US_Elec[-1]
VAR_est <- VAR(cbind(Coal_Futures_returns, Coal_US_Elec), ic="AIC", lag.max = 24)
coeftest(VAR_est)
causality(VAR_est, cause="Coal_Futures_returns")["Granger"] # Testing for Granger-causal effects of the Coal_US_Elec on Coal futures returns
causality(VAR_est, cause="Coal_US_Elec")["Granger"] # Testing for Granger-causal effects of coal futures returns on Coal_US_Elec

# Interpretation
# TBD

# Degression: Plotting impulse response functions
plot(irf(VAR_est, impulse="Coal_US_Elec", response="Coal_Futures_returns")) # We shock Coal_Futures_price_diff at time t
plot(irf(VAR_est, impulse="Coal_Futures_returns", response="Coal_US_Elec")) # We shock Coal_US_Elec at time t
# Interpretation:


```

# GARCH model to be developed