---
title: "Investigation of the relationship between price of futures contracts of energy commodities and their energy generation contribution to US energy grid"
author: "Luca Renz & Kasper Aleksander Kuznik"
date: "December 2024"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
    theme: simplex
---

```{r setup, include=FALSE}
# Global options
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```

\newpage

# 1. Introduction
Energy generation and its relationship to commodity markets have significant implications for financial stability and market behavior. This paper examines the influence of U.S. energy generation—categorized by sources such as natural gas, crude oil, and coal—on the corresponding commodities futures prices. By employing time series analysis, we explore how shifts in energy generation impact the pricing dynamics of these key commodities. While the focus of this documentation is on the primary findings and visualizations, supplementary charts and detailed analyses are included in the appendix.

# 2. Motivation
The interplay between commodity markets and energy generation is pivotal to understanding economic resilience and environmental sustainability in the contemporary energy landscape. This study is motivated by the need to dissect the dynamics between U.S. energy production and the futures prices of critical commodities—namely natural gas, crude oil, and coal. These commodities not only fuel the nation but also significantly influence the financial markets due to their integral role in energy generation. Therefore, the hypothesis under test in this study is that changes in U.S. energy generation impact the futures prices of natural gas, crude oil, and coal. It is aimed to verify this by analyzing time series data to understand if energy production fluctuations can predictably influence market trends.

# 3. Methods
To investigate the causal relationship between U.S. energy generation by source and commodities futures prices for Natural Gas, Crude Oil, and Coal, the first step is to prepare the data to ensure comparability. After cleansing and preprocessing the data, the "Augmented Dickey-Fuller Test" is applied to determine whether the time series are stationary. If any time series are found to be non-stationary, they are transformed using the diff(log()) method until the "Dickey-Fuller Test" confirms stationarity. Once the series are stationary, they are used to construct vector autoregressive models (VAR models) to analyze the dynamic relationships. Additionally, Granger causality tests are performed to evaluate the direction and strength of causality between the variables.

## 3.1 Data collection and preprocessing
The historical futures prices for Natural Gas, Crude Oil (WTI), and Newcastle Coal were sourced from Barchart.com in the form of CSV files. Data on U.S. electricity generation, categorized by energy source, was obtained from the U.S. Energy Information Administration as XLSX file. This method required manual extraction and preparation. The availability of data varies: Crude Oil futures pricing data begins in 1983, Natural Gas futures data in 1990, Coal futures data in 2009, and U.S. energy generation data in 2001. All datasets are reported on a monthly basis and are available up to August 2024. Despite these differences in starting dates, the data was aligned to ensure consistency for the analysis period.
```{r libraries, results = "hide"}
# Load necessary libraries
#Install and import libraries.
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("lubridate")) install.packages("lubridate")
if (!require("vars")) install.packages("vars")
if (!require("forecast")) install.packages("forecast")
if (!require("rugarch")) install.packages("rugarch")
if (!require("dplyr")) install.packages("dplyr")
if (!require("purrr")) install.packages("purrr")
if (!require("readxl")) install.packages("readxl")
if (!require("janitor")) install.packages("janitor")
if (!require("tseries")) install.packages("tseries")

library(tidyverse)
library(dplyr)
library(lubridate)
library(vars)
library(forecast)
library(rugarch)
library(purrr)
library(readxl)
library(janitor)  # For cleaning column names
library(tseries) # For adf.test()
library(quantmod) # - this package is used to fetch marcoeconimic data from FED
```

```{r data-prep, results = "hide", cache=TRUE}
coal_futures <- read.csv("data/NewCastle_Coal_Futures_historical_prices.csv")
elec_generation <- readxl::read_excel("data/US_Elec_Generation_Monthly.xlsx")
all_sheets <- excel_sheets("data/US_Elec_Generation_Monthly.xlsx")
oil_futures <- read.csv("data/Crude_Oil_WTI_Futures_historical_prices.csv")
gas_futures <- read.csv("data/Natural_Gas_Futures_historical_prices.csv")

# Natural Gas Futures
gas_futures <- gas_futures %>%
    mutate(YearMonth = floor_date(as.Date(Exp.Date, format = "%d/%m/%Y"), "month")) %>%
  dplyr::select(YearMonth, Last) %>%
  rename(Gas_Price = Last)

# Oil Futures
oil_futures <- oil_futures %>%
    mutate(YearMonth = floor_date(as.Date(Exp.Date, format = "%Y-%m-%d"), "month")) %>%
  dplyr::select(YearMonth, Last) %>%
  rename(Oil_Price = Last)

# Coal Futures
coal_futures <- coal_futures %>%
  mutate(YearMonth = floor_date(as.Date(Exp.Date, format = "%Y-%m-%d"), "month")) %>%
  dplyr::select(YearMonth, Last) %>%
  rename(Coal_Price = Last)

# Function to remove all single quotes from character columns
remove_quotes <- function(df) {
  df %>%
    mutate(across(where(is.character), ~ gsub("'", "", .))) %>%
     mutate(across(where(is.character), ~ gsub("\\n|\\r", " ", .)))  # Remove line breaks
}

valid_sources <- c("Coal", "Gas", "Oil")
# not as wide table
elec_generation_processed <- all_sheets %>%
  lapply(function(sheet) {
    read_excel("data/US_Elec_Generation_Monthly.xlsx", sheet = sheet) %>%
      remove_quotes() %>%
      clean_names() %>%
      mutate(
  energy_source = case_when(
    energy_source == "Natural gas" ~ "Gas",
    energy_source == "Other Gases" ~ "Gas",
    energy_source == "Petroleum" ~ "Oil",
    TRUE ~ energy_source
  )
) %>%
     filter(state == "US-Total") %>% 
      filter(startsWith(type_of_producer, "Total")) %>% 
     filter(energy_source %in% valid_sources) %>%
      mutate(
        year_month = make_date(year, month)  # Create a YearMonth column
      ) %>%
      dplyr::select(year_month, energy_source, mwh) %>%
      rename(YearMonth = year_month)
  }) %>%
  bind_rows()  # Combine all processed data from different sheets

sum(is.na(coal_futures$YearMonth))
sum(is.na(gas_futures$YearMonth))
sum(is.na(oil_futures$YearMonth))


wide_table <- elec_generation_processed %>%
  pivot_wider(
    names_from = energy_source,  # Use 'energy_source' for column names
    values_from = mwh,          # Use 'mwh' for the values in those columns, fill missing values with 0
  )


# Merge all datasets
merged_data <- purrr::reduce(
  list(coal_futures, oil_futures, gas_futures, wide_table),
  dplyr::left_join,
  by = "YearMonth"
)


write.csv(merged_data,'data/merged_data.csv', row.names=FALSE)
```

## 3.2 Data Preparation and Visualization
```{r analysis, cache=TRUE}
# Data filtering from 2012-01-01 to 2024-08-01
merged_data_reduced <- merged_data[38:189,]
# Futures prices
Nat_Gas_Futures_price <- ts(merged_data_reduced$Gas_Price, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
Oil_Futures_price <- ts(merged_data_reduced$Oil_Price, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
Coal_Futures_price <- ts(merged_data_reduced$Coal_Price, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
# US Electricity
Nat_Gas_US_Elec <- ts(merged_data_reduced$Gas, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
Oil_US_Elec <- ts(merged_data_reduced$Oil, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)
Coal_US_Elec <- ts(merged_data_reduced$Coal, start = as.yearmon(merged_data_reduced$YearMonth[1]), frequency = 12)

# Plotting the data
# Futures data
par(mfrow = c(2,3), mar = c(4, 4, 2, 1), oma = c(1, 1, 3, 1), cex.main = 0.8, cex.axis = 0.7, cex.lab = 0.8)
plot(Nat_Gas_Futures_price, xlab = "Time", ylab = "Natural Gas Futures Price [tsd USD]", main = "Natural Gas Future Price\n Jan2012 - Aug2024")
plot(Oil_Futures_price, xlab = "Time", ylab = "Oil Futures Price [USD]", main = "Oil Future Price\n Jan2012 - Aug2024")
plot(Coal_Futures_price, xlab = "Time", ylab = "Coal Futures Price [USD]", main = "Coal Future Price\n Jan2012 - Aug2024")
# US Electricity data
plot(Nat_Gas_US_Elec, xlab = "Time", ylab = "US Electricity Generation - Natural Gas", main = "Natural GAs US Electricity Generation\n Jan2012 - Aug2024")
plot(Oil_US_Elec, xlab = "Time", ylab = "US Electricity Generation - Oil", main = "Oil US Electricity Generation\n Jan2012 - Aug2024")
plot(Coal_US_Elec, xlab = "Time", ylab = "US Electricity Generation - Coal", main = "Coal US Electricity Generation\n Jan2012 - Aug2024")
```

## 3.3 Time Series Sationarity
To develop the VAR models, it is essential to first ensure that all time series are stationary. A stationary time series is defined as one with consistent statistical properties over time, such as mean, variance, and autocorrelation. To verify stationarity, the Augmented Dickey-Fuller test is applied to each of the investigated time series. The test results, including p-values, are used to assess the stationarity of the time series data, such as futures prices of Natural Gas, Crude Oil, Coal, as well as U.S. energy generation by the energy source. 

```{r, results='hide'}
# Testing for stationarity for both time series and creating stationary time series of same lengths
# Futures data
adf.test(Nat_Gas_Futures_price) # Test for stationarity of the Natural Gas Future price series 
adf.test(Oil_Futures_price) # Test for stationarity of the Oil Futures price series
adf.test(Coal_Futures_price) # Test for stationarity of the Coal Futures price series

# US Electricity data
adf.test(Nat_Gas_US_Elec) # Test for stationarity of Nat Gas US Elec Gen Monthly series
adf.test(Oil_US_Elec) # Test for stationarity of Oil US Elec Gen Monthly series
adf.test(Coal_US_Elec) # Test for stationarity of Coal US Elec Monthly generation series
```

| Time series                           | p-value |
|---------------------------------------|---------|
| Natural Gas Futures Price             |  0.353  |
| Crude Oil Futures Price               |  0.610  |
| Coal Futures Prices                   |  0.056  |
| US Electricity Generation Natural Gas |  0.068  |
| US Electricity Generation Crude Oil   |  0.010  |
| US Electricity Generation Coal        |  0.010  |

*Table 1: Augmented Diickey-Fuller test results for investigated time series*

Time series with a P-value greater than 0.05, such as Natural Gas U.S. energy generation and futures prices of Natural Gas, Crude Oil, and Coal, are classified as non-stationary. To address this, stationarity is introduced by applying diff(log()) transformation.

```{r, results='hide'}
# Apply diff-log-transformation if necessary
# Futures Data
Nat_Gas_Futures_returns <- diff(log(Nat_Gas_Futures_price))
Oil_Futures_returns <- diff(log(Oil_Futures_price))
Coal_Futures_returns <- diff(log(Coal_Futures_price))

# US Electricity Data
Nat_Gas_US_Elec_diff <- diff(log(Nat_Gas_US_Elec))

# Plotting the differences
par(mfrow = c(2,3), mar = c(4, 4, 2, 1), oma = c(1, 1, 3, 1), cex.main = 0.8, cex.axis = 0.7, cex.lab = 0.8)
plot(Nat_Gas_Futures_returns, xlab = "Time", ylab = "Natural Gas Futures Returns", main = "Natural Gas Future Returns\n Jan2012 - Aug2024")
plot(Oil_Futures_returns, xlab = "Time", ylab = "Oil Futures Returns", main = "Oil Gas Future Returns\n Jan2012 - Aug2024")
plot(Coal_Futures_returns, xlab = "Time", ylab = "Coal Futures Returns", main = "Coal Gas Future Returns\n Jan2012 - Aug2024")
plot(Nat_Gas_US_Elec_diff, xlab = "Time", ylab = "Natural Gas US Electricity Generation first difference", main = "Natural Gas US Electricity Generation First Difference\n Jan2012 - Aug2024")

# Check for stationarity again
adf.test(Nat_Gas_Futures_returns) # Test for stationarity of Diff of Nat Gas US Elec Gen Monthly series
adf.test(Oil_Futures_returns) # Test for stationarity of Diff of Oil US Elec Gen Monthly series
adf.test(Coal_Futures_returns) # Test for stationarity of Diff of Coal US Elec Monthly generation series
adf.test(Nat_Gas_US_Elec_diff) # Test for stationarity of Diff of Coal US Elec Monthly generation series
```

| Time series                           | p-value |
|---------------------------------------|---------|
| Natural Gas Futures Price             |   0.01  |
| Crude Oil Futures Price               |   0.01  |
| Coal Futures Prices                   |   0.01  |
| US Electricity Generation Natural Gas |   0.01  |

*Table 2: Augmented Dickey-Fuller test results for investigated time series after first differences*

After applying the diff(log()), the previously insignificant results (non-stationary) have been transformed to significant results below the threshold of 0.05.

# 4. VAR Model and Granger Causality
Once the time series are made stationary by applying first differences, the next step is to fit a Vector Autoregressive model to the stationary data. The VAR model captures the linear relationships between a variable and its own past values, as well as the past values of other endogenous variables. This approach allows for the exploration of the dynamics and feedback among multiple interrelated time series, offering insights into the causal relationships within the data. 

## 4.1a Natural Gas Futures Returns vs Natural Gas US Electricity Generation Difference
In this analysis, the relationship between differenced natural gas futures returns and changes in U.S. electricity generation from natural gas was examined using VAR models. The Granger causality tests indicated no significant predictive relationship. Detailed results and plots are available in the appendix.

## 4.1b Natural Gas Futures Returns vs Natural Gas US Electricity Generation
Natural Gas Data is not stationary but give p-value 0.06 checking the results out of curiosity
This section explored the relationship between natural gas futures returns and U.S. electricity generation from natural gas without differencing. The results were insignificant, underscoring the importance of using differenced data for stationarity. Exact results and plots can be found in the appendix.


## 4.2 Oil Futures Returns vs Oil US Electricity Generation
The relationship between oil futures returns and U.S. electricity generation from oil was analyzed. Granger causality tests showed no significant effects, confirming that changes in oil-based electricity generation do not predict oil futures prices. Detailed results and plots are available in the appendix.

## 4.3 Coal Futures Returns vs Coal US Electricity Generation
This analysis investigated the impact of changes in coal-based U.S. electricity generation on coal futures returns. The results were insignificant, indicating no predictive causality between the two. Exact results and plots can be found in the appendix.


# 5. Results & Conclusion
The study initiated with Augmented Dickey-Fuller tests which identified non-stationarity in the time series data concerning U.S. energy production and commodity futures prices. To achieve stationarity, necessary for accurate econometric analysis, a diff(log()) transformation was applied. This method effectively stabilized the variance and mean of the series, allowing for further analysis with Vector Autoregression (VAR) models.

Employing VAR models and subsequent Granger causality tests, the study aimed to determine if historical data on energy production could predict futures prices for natural gas, crude oil, and coal. The causality tests, however, did not reveal significant predictive relationships between these variables.

**These findings from section 4 suggest that fluctuations in U.S. energy production do not significantly influence the futures prices of these commodities**, indicating that other external factors such as global economic conditions, market sentiment, or geopolitical events likely have a greater impact. This highlights the complexity of the factors driving commodity markets and suggests that a broader perspective is necessary for understanding and forecasting movements in these prices.

# 6. Appendix
## Section 4 - Additional information

### 4.1a Natural Gas Futures Returns vs Natural Gas US Electricity Generation Difference
```{r, results='hide', fig.width=6, fig.height=4}
# Estimating vector autoregession and testing for causality
VAR_est <- VAR(cbind(Nat_Gas_Futures_returns, Nat_Gas_US_Elec_diff), ic="AIC", lag.max = 24)
coeftest(VAR_est)
causality(VAR_est, cause="Nat_Gas_Futures_returns")["Granger"] # Testing for Granger-causal effects of Natural Gas US Electricity Generation on Natural Gas Futures Returns
causality(VAR_est, cause="Nat_Gas_US_Elec_diff")["Granger"] # Testing for Granger-causal effects of Natural Gas Futures Returns on Natural Gas US Electricity Generation

# Plotting impulse response functions
par(mfrow = c(1,2), mar = c(4, 4, 2, 1), oma = c(1, 1, 3, 1), cex.main = 0.8, cex.axis = 0.7, cex.lab = 0.8)
plot(irf(VAR_est, impulse="Nat_Gas_US_Elec_diff", response="Nat_Gas_Futures_returns"), xlab = "Lag", ylab = "Natural Gas Futures Returns", main = "Orthogonal Impulse Response \nfrom Natural Gas US Electricity Generation Diff") # We shock Nat_Gas_Futures_price_diff at time t
plot(irf(VAR_est, impulse="Nat_Gas_Futures_returns", response="Nat_Gas_US_Elec_diff"), xlab = "Lag", ylab = "Natural Gas US Electricity Generation Diff", main = "Orthogonal Impulse Response \nfrom Natural Gas Futures Returns") # We shock Nat_Gas_US_Elec_Diff at time t
```

### 4.1b Natural Gas Futures Returns vs Natural Gas US Electricity Generation

```{r, results='hide', fig.width=6, fig.height=4}
# Estimating vector autoregession and testing for causality
Nat_Gas_US_Elec <- Nat_Gas_US_Elec[-1]
VAR_est <- VAR(cbind(Nat_Gas_Futures_returns, Nat_Gas_US_Elec), ic="AIC", lag.max = 24)
coeftest(VAR_est)
causality(VAR_est, cause="Nat_Gas_Futures_returns")["Granger"] # Testing for Granger-causal effects of Natural Gas US Electricity Generation on Natural Gas Futures Returns
causality(VAR_est, cause="Nat_Gas_US_Elec")["Granger"] # Testing for Granger-causal effects of Natural Gas Futures Returns on Natural Gas US Electricity Generation

# Plotting impulse response functions
par(mfrow = c(1,2), mar = c(4, 4, 2, 1), oma = c(1, 1, 3, 1), cex.main = 0.8, cex.axis = 0.7, cex.lab = 0.8)
plot(irf(VAR_est, impulse="Nat_Gas_US_Elec", response="Nat_Gas_Futures_returns"), xlab = "Lag", ylab = "Natural Gas Futures Returns", main = "Orthogonal Impulse Response from \nNatural Gas US Electricity Generation") # We shock Nat_Gas_Futures_price_diff at time t
plot(irf(VAR_est, impulse="Nat_Gas_Futures_returns", response="Nat_Gas_US_Elec"), xlab = "Lag", ylab = "Natural Gas US Electricity Generation", main = "Orthogonal Impulse Response \nfrom Natural Gas Futures Returns") # We shock Nat_Gas_US_Elec at time t
```

### 4.2 Oil Futures Returns vs Oil US Electricity Generation

```{r, results='hide', fig.width=6, fig.height=4}
Oil_US_Elec <- Oil_US_Elec[-1]
VAR_est <- VAR(cbind(Oil_Futures_returns, Oil_US_Elec), ic="AIC", lag.max = 24)
coeftest(VAR_est)
causality(VAR_est, cause="Oil_Futures_returns")["Granger"] # Testing for Granger-causal effects of the US Oil Electricity Generation on oil futures returns
causality(VAR_est, cause="Oil_US_Elec")["Granger"] # Testing for Granger-causal effects of oil futures returns on US Oil Electricity Generation

# Plotting impulse response functions
par(mfrow = c(1,2), mar = c(4, 4, 2, 1), oma = c(1, 1, 3, 1), cex.main = 0.8, cex.axis = 0.7, cex.lab = 0.8)
plot(irf(VAR_est, impulse="Oil_US_Elec", response="Oil_Futures_returns"), xlab = "Lag", ylab = "Oil Futures Returns", main = "Orthogonal Impulse Response \nfrom Oil US Electricity Generation") # We shock Oil futures returns at time t
plot(irf(VAR_est, impulse="Oil_Futures_returns", response="Oil_US_Elec"), xlab = "Lag", ylab = "Oil US Electricity Generation", main = "Orthogonal Impulse Response \nfrom Oil Futures Returns") # We shock US Oil Electricity Generation at time t
```

### 4.3 Coal Futures Returns vs Coal US Electricity Generation

```{r, results = 'hide', fig.width=6, fig.height=4}
Coal_US_Elec <- Coal_US_Elec[-1]
VAR_est <- VAR(cbind(Coal_Futures_returns, Coal_US_Elec), ic="AIC", lag.max = 24)
coeftest(VAR_est)
causality(VAR_est, cause="Coal_Futures_returns")["Granger"] # Testing for Granger-causal effects of the Coal_US_Elec on Coal futures returns
causality(VAR_est, cause="Coal_US_Elec")["Granger"] # Testing for Granger-causal effects of coal futures returns on Coal_US_Elec

# Plotting impulse response functions
par(mfrow = c(1,2), mar = c(4, 4, 2, 1), oma = c(1, 1, 3, 1), cex.main = 0.8, cex.axis = 0.7, cex.lab = 0.8)
plot(irf(VAR_est, impulse="Coal_US_Elec", response="Coal_Futures_returns"), xlab = "Lag", ylab = "Coal Futures Returns", main = "Orthogonal Impulse Response from \nCoal US Electricity Generation") # We shock Coal_Futures_price_diff at time t
plot(irf(VAR_est, impulse="Coal_Futures_returns", response="Coal_US_Elec"), xlab = "Lag", ylab = "Coal US Electricity Generation", main = "Orthogonal Impulse Response \nfrom Coal Futures Returns") # We shock Coal_US_Elec at time t
```


